### 2020-06-22 chapters/background.tex
EMA-like histogram

\textcolor{red}{Sliding window Weights image}
\textcolor{red}{EMA Weights image}

@@

\begin{definition}
EMA\textsubscript{t} = EMA\textsubscript{t-1} * $\lambda$ + val\textsubscript{t}

\label{def:ema}
\end{definition}

with $\lambda$ as a constant factor applied to the previous EMA value. This $\lambda$ constant (0 < $\lambda \leq$ 1) is known as the smoothing coefficient. The smoothing or decay factor $\lambda$ is given by:

\[ \lambda = 2^{- \dfrac{1}{\tau}} \]

where $\tau$ is the half-life of the EMA. The half-life $\tau$ controls how fast older events are forgotten as the smoothing factor depends on it. The half-life is the amount of time that it takes for an element to see its contribution or weight reduced to half. For instance, with $\tau = 1s$ we have $\lambda = 0.5$, which means for each 1s step we reduce the weights applied to each event by half. Note that it is also possible to express a tuple-based half-life, \textit{i.e.}, how many tuples or events must the EMA process before a certain element has its weight reduced by half (\textit{e.g.} $\tau = 1$ tuple, $\lambda = 0.5$, for each new tuple reduce all weights by half).

% EMA-count example (later build histogram concept on multiple EMA-counts)

>>>>

\textcolor{red}{TODO: Images for EMA and sliding window weights}

@@

\begin{definition}
EMA\textsubscript{t} = EMA\textsubscript{t-1} * $\lambda\delta\textsubscript{t}$ + val\textsubscript{t}
\label{def:ema}
\end{definition}

with $\lambda$ as a constant factor applied to the previous EMA value and $\delta\textsubscript{t}$ as the time passed between \textit{t} and \textit{t-1}. This $\lambda$ constant (0 < $\lambda \leq$ 1) is known as the smoothing coefficient. The smoothing or decay factor $\lambda$ is given by:
\begin{equation}
    \lambda = 2^{- \dfrac{1}{\tau}}
    \label{eq:ema-decay}
\end{equation}
where $\tau$ is the half-life of the EMA. The half-life $\tau$ controls how fast old events are forgotten as the smoothing factor depends on it. The half-life is the amount of time that it takes for an element to see its contribution or weight reduced to half. For instance, with $\tau = 1s$ we have $\lambda = 0.5$, which means for each 1s step we reduce the weights applied to each event by half.

Note that it is also possible to express a tuple-based half-life, \textit{i.e.}, how many tuples or events must the EMA process before a certain element has its weight reduced by half. In this case, the definition does not take into account the time passed between events:
\begin{definition}
EMA\textsubscript{t} = EMA\textsubscript{t-1} * $\lambda$ + val\textsubscript{t}
\label{def:tuple-ema}
\end{definition}
For example, with a tuple-based half-life of $\tau = 1$ tuple, the smoothing factor is $\lambda = 0.5$, meaning for each new tuple all weights are reduced by half.

In this Thesis we will work with tuple-based EMAs but our approach can be adjusted to use time-based ones.

### 2020-06-22 chapters/my-work.tex
EMA-like histogram

\section{Approach A: Usage of outlier detection methods}
The first approach considered relied on outlier detection methods such as the ones presented in Section \ref{sec:outliers} and classified by the taxonomy in Figure \ref{fig:outlier-taxonomy}. This straightforward approach works with either subsequence or point outlier detection methods to analyze the streaming sliding window and produce data pattern shift alerts.

@@

Using point outlier detection methods, the main idea is to keep track of the rate of point outliers detected. For instance, keep track of how many outliers were alerted in the past hour. The decision to alert for a data pattern shift of the system would be made when comparing this rate of point outliers detected in the past hour with a user-defined threshold $\alpha$. On the other hand, using subsequence outlier detection methods would allow us to report the whole sliding window as an outlier.

In our lightweight real-time solution we need constant time complexity to process events one by one in real-time. Because our system is not mission-critical, we also want the solution to be as lightweight as possible, which means techniques that grow linearly with window size are not a good fit. Therefore, most of the outlier detection methods presented become unfit for our use due to high memory consumption and time complexity. Some of the methods discussed also suffer from what we call \textit{"memory loss"}: they lose track of the reference window or normality period. In other words, the algorithms are online learners, meaning they adapt to the underlying statistics of the data stream. While in some scenarios this is a much-needed feature, in our use-case it is not. How clear the alerts and insights retrieved from the methods are, is also an important property. Some outlier detection methods use Machine Learning (ML) techniques to produce alerts, which makes it harder to explain to a system administrator why an alert was raised and what changed.

We concluded that employing outlier detection methods was not the best course of action and put aside this approach.

\section{Approach B: Thresholding aggregation values}
This second approach focus on building an aggregation snapshot of the reference period to compare with streaming aggregation values and alert if the difference between both surpasses a threshold $\alpha$.

In this approach we compute sliding window aggregations and compare the aggregations’ current state versus an initial reference, reporting a shift when the difference is substantial. For instance, consider the case where we monitor one single feature or variable \textit{x}. Assume we define the set of aggregations to compute the mean of \textit{x} and its standard deviation. Also assume we save the reference window and maintain a sliding window and the average and standard deviation aggregations for \textit{x} efficiently. As seen in Figure \ref{fig:approach2-initial-state}, the initial mean and standard deviation values for \textit{x} presented in the reference window are of 5 and 1, respectively. Later on, in \textit{Window 1}, we see the same values for the mean and standard deviation, of 5 and 1, respectively. This represents a \textit{normal} case where no alert is produced.

@@

\subsection*{Challenges}
The first issue with this approach is that generic sliding window aggregation algorithms grow linearly in space regarding window size, such as Recalculate-From-Scratch and Subtract-On-Evict presented in Section \ref{sec:back-swag-algs}, Two-Stacks has seen in Section \ref{sec:2stacks} and DABA in Section \ref{sec:daba}. Another challenge when using this approach is defining the set of aggregations to compute at feature/variable level that represent the system state well enough for comparison of reference and streaming periods and change detection.

@@

\section{Proposed method: Feature Distribution Monitoring} \label{sec:ft-monitoring}

@@

We do not need to directly compare the entire window contents of the reference and sliding window periods. The latter would require storing all window contents hence making the memory consumption grow linearly with sliding window and/or reference period size. Instead, we want to use a  \textit{(a)} constant in memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase. Additionally, we want a streaming aggregation that \textit{(c)} reliably encodes the distribution of each feature.

@@

In Section \ref{sec:emas} we discussed Exponential Moving Averages (EMAs) and how they emulate a sliding window without actually storing window elements....

\textcolor{red}{TODO: complete this after EMAs in background are corrected}

>>>>

\section{State of the art outlier detection methods}
The first approach considered relied on outlier detection methods such as the ones presented in Section \ref{sec:outliers} and classified by the taxonomy in Figure \ref{fig:outlier-taxonomy}.

We reviewed subsequence and point outlier detection methods. Using point outlier detection methods, the main idea is to keep track of the rate of point outliers detected. For instance, keep track of how many outliers were alerted in the past hour. The decision to alert for a data pattern shift of the system would be made when comparing this rate of point outliers detected in the past hour with a user-defined threshold $\alpha$. On the other hand, using subsequence outlier detection methods would allow us to report the whole sliding window as an outlier when compared to the reference window.

For this Thesis we need constant time complexity to process events one by one in real-time. Because our system is not mission-critical, we also want the solution to be as lightweight as possible, which means techniques that grow linearly with window size are not a good fit. Some of the methods discussed also suffer from what we call \textit{"memory loss"}: they lose track of the reference window or normality period. In other words, the algorithms are online learners, meaning they adapt to the underlying statistics of the data stream. While in some scenarios this is a much-needed feature, in our use-case it is not. How clear the alerts and insights retrieved from the methods are, is also an important property. Some outlier detection methods use Machine Learning (ML) techniques to produce alerts, which makes it harder to explain to a system administrator why an alert was raised and what changed.

\section{State of the art aggregation algorithms}
This second approach focus on building an aggregation snapshot of the reference period to compare with streaming aggregation values and alert if the difference between both surpasses a threshold $\alpha$.

@@

In this approach we compute sliding window aggregations and compare the aggregations’ current state versus an initial reference, reporting a shift when the difference is substantial. For instance, consider the case where we monitor one single feature or variable \textit{x}. Assume we define the set of aggregations to compute the mean of \textit{x} and its standard deviation. Also assume we save the reference window and maintain a sliding target window and the associated average and standard deviation aggregations for \textit{x} efficiently. As seen in Figure \ref{fig:approach2-initial-state}, the initial mean and standard deviation values for \textit{x} presented in the reference window are of 5 and 1, respectively. Later on, in \textit{Window 1}, we see the same values for the mean and standard deviation, of 5 and 1, respectively. This represents a \textit{normal} case where no alert is produced.

@@

\subsection*{Challenges}
The first issue with this approach is that generic sliding window aggregation algorithms grow linearly in space regarding window size, such as Recalculate-From-Scratch and Subtract-On-Evict from Section \ref{sec:back-swag-algs}, Two-Stacks from Section \ref{sec:2stacks} and DABA in Section \ref{sec:daba}. Another challenge when using this approach is defining the set of aggregations to compute at feature/variable level that represent the system state well enough for comparison of reference and streaming periods and change detection.

@@

\section{Method: Feature Distribution Monitoring} \label{sec:ft-monitoring}

@@

We do not need to directly compare the entire window contents of the reference and sliding windows. The latter would require storing all window contents hence making the memory consumption grow linearly with sliding window and/or reference period size. Instead, we want to use a  \textit{(a)} constant in memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase. Additionally, we want a streaming aggregation that \textit{(c)} reliably encodes the distribution of each feature.

@@

In Section \ref{sec:emas} we discussed Exponential Moving Averages (EMAs) and how they compute sliding window aggregations without actually storing the window contents. Definition \ref{def:tuple-ema} shows how to compute an EMA using its recurrence formula. An EMA state or value depends only on the previous one times an exponential decay $\lambda$. Recall that this decay is given by Equation \ref{eq:ema-decay}:
\[ \lambda = 2^{- \dfrac{1}{\tau}} \]
where $\tau$ is commonly denoted as the half-life.

After processing \textit{x} tuples the oldest tuple is \emph{smoothed} by a factor of $\lambda^x = 2^{- \dfrac{x}{\tau}}$. If we process $x = 4\tau$ tuples the oldest tuple would have its contribution reduced by
\[ 2^{- \dfrac{4\tau}{\tau}} = 2^{-4} \]
which is approximately 0.06. In other words, the \textit{x-th} event would have 6\% contribution towards the EMA state or value meaning older events would have even less weight.

We choose to use windows with four times the half-life size (or $4\tau$), meaning we discard events whose contribution is less than 6\%. In fact, we set the half-life to be one quarter of the size of the window we want to emulate. For instance, if we want to compute an EMA aggregation over a period of a month, we set the half-life to be the number of tuples that correspond to one week worth of data.

### 2020-06-22 chapters/sota.tex
EMA-like histogram

Density-based univariate point outlier detection methods rely on the clustering of data points and neighbor counting. These methods consider a datum an outlier if it has less than $\alpha$ neighbors. Distance between two points is usually measured using the Euclidean distance. Two points are said to be neighbors if they are close to each other --- \textit{i.e.} the distance between them is less than or equal to a given threshold $\alpha$.

@@

\subsubsection*{Applicability to our Hypothesis}
In this Thesis, we want to detect and alert data pattern shifts with a low memory footprint. However, the method proposed by Kifer et. al grows linearly with window size, making it unfit for large windows, our use-case.

>>>>

Density-based univariate point outlier detection methods rely on the clustering of data points and neighbor counting. These methods consider a datum an outlier if it has less than $\alpha$ neighbors. Distance between two points is usually measured using the Euclidean distance \cite{elementary-algebra}. Two points are said to be neighbors if they are close to each other --- \textit{i.e.} the distance between them is less than or equal to a given threshold $\alpha$.

@@

\subsubsection*{Applicability to our Hypothesis}
In this Thesis, we want to detect and alert data pattern shifts with a low memory footprint. The method proposed by Kifer et. al grows linearly with window size, making it unfit for large windows, our use-case.

However, this two-window paradigm encodes a static normal or reference period in one window and the observed streaming values in a sliding one. Since in this Thesis we want to report streaming data pattern shifts relatively to a reference period, we may use a similar two-windowed model if we find lightweight and incrementally maintained aggregations over both windows.

\subsubsection{Automatic Model Monitoring for Data Streams}

ALSO a 2 WINDOW MODEL
\textcolor{red}{TODO: write this}
